# R cheatsheet on data transforamtion and exploration

Sai Krupa Jangala

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Few Pointers:

* The purpose of this cheatsheet is to describe all the basic data operations that we do when we start a new project. This also includes different types of data transformations, explorations and management.
* We use only data frames throughtout the cheatsheet.
* We use packages that are most commonly used in R. We will be using these packages for different purposes. 
* Different datasets are used to best illustrate the transformation, management and exploration.
* The data sets used are ```openintro::fastfood``` , ```openintro::seatlepets```, ```cars```, ```openintro::ames``` and ```mtcars```.
* Every chunk of code loads its own data, if it's not, it means that we have used the data from the previous chunk.

#### Required packages
These packages are most commonly in R and are used in the following cheatsheet. We will be using these packages for different purposes. 
```{r, message=FALSE}
#install.packages("tidyverse")
#install.packages("dplyr")
#install.packages("reshape")
library(dplyr)
library(tidyverse)
library(reshape)
```

#### Output the head, tail and sample of the dataframe.
Head - will get the first 5 rows of the dataframe.
```{r}
data <- cars
head(data, n=5)
```
Tail - will get the last 5 rows of the dataframe
```{r}
tail(data, n=5)
```

#### Selection of only a few columns from a dataframe.
Select only restaurant, item and calorie columns from the fastfood dataset. Two variations are shown below.
```{r}
data <- openintro::fastfood
# Variation 1
data <- data[, c("restaurant","item","calories")]
# Variation 2
data_1 <- select(data, c(restaurant, item, calories))
```

As we can see below, both the transformations produce the same result.
```{r}
head(data)
```
```{r}
head(data_1)
```
#### Get the column names
Generate all the column names in the data frame
```{r}
data <- openintro::fastfood
names(data)
```

#### Drop one or more columns.
Drop columns ```restaurant```, ```item```, ```calories``` from the data. As we can observe, the above column names have been dropped.
```{r}
data <- openintro::fastfood
data <- select(data, -c(restaurant, item, calories))
names(data)
```

#### Transformation using the transform() function in R

Here we are creating a new dataframe by changing the speed column. We transformed it by multiplying with 100
```{r}
data <- cars
data_1 <- transform(data, speed=speed*100)
head(data_1)
```
Here we transformed the original dataframe by creating a new column.
```{r}
data <- transform(data, time=speed*dist)
head(data)
```
#### Conditional Transformation in R

Transformation based on a condition.  Here we are creating a new column called Grilled, which is assigned to Grilled if the item contains Grilled in it's name else it is classified as Not Grilled.
```{r}
#Check if it contains the word grilled.
data <- openintro::fastfood
data <- transform(data, Grilled=ifelse(str_detect(item, "Grilled"), "Grilled", "Not Grilled"))
head(data[,c("item","Grilled")])
```
#### Add a new column to the dataframe without ```transform()``` function.

Here we added a new column called time.
```{r}
data <- cars
data$time <- data$speed * data$dist
head(data)
```

#### Get all the unique values of a column in a dataframe.
```{r}
data <- openintro::fastfood
unique(data$restaurant)
```

#### Using the filter function in R. Different logical operators can be used to filter the data.

Filter all the rows where mpg column value is 21.0
```{r}
data <- mtcars
data_filtered <- filter(data, mpg==21.0)
unique(data_filtered$mpg)
```

Filter all the rows where mpg column value is less than 21.0
```{r}
data_filtered_1 <- filter(data, mpg<21.0)
unique(data_filtered_1$mpg)
```
Filter all the rows where mpg column value is greater than 21.0
```{r}
data_filtered_2 <- filter(data, mpg>21.0)
unique(data_filtered_2$mpg)
```
Filter all the rows where cyl column is 4 and carb column is greater than 1.
```{r}
data_filtered_logical <- filter(data, cyl == 4 & carb > 1)
unique(data_filtered_logical$mpg)
```

#### Select only few rows in a column based on a condition without using the ```filter()``` function.

Here we are selecting only those rows where the restuarant name is subway
```{r}
data <- openintro::fastfood
# Select the rows where the item contains the word "Grilled
data <- data[data$restaurant == "Subway", ] 
unique(data$restaurant)
```

#### Merge two dataframes

Merging two dataframes based on column names. Authors dataframe and books dataframe have been merged by surname and name.

```{r}
authors <- data.frame(
    surname = c("Tukey", "Venables", "Tierney", "Ripley", "McNeil"),
    nationality = c("US", "Australia", "US", "UK", "Australia"),
    retired = c("yes", rep("no", 4)))
books <- data.frame(
    name = c("Tukey", "Venables", "Tierney", "Ripley", "Ripley", "McNeil"),
    title = c("Exploratory Data Analysis",
              "Probability and Statistics",
              "Finance and Structuring for Data Science",
              "Algorithms for Data Science",
               "Interactive Data Analysis",
              "Deep Learning"))
    #other.author = c(NA, "Ripley", NA, NA, NA, NA))
merged <- merge(authors, books, by.x="surname", by.y="name")
head(merged)
```


#### Arrange the data in ascending order based on a column
```{r}
data <- openintro::fastfood
# Arranging the data in ascending order
data <- data[order(data$total_fat),] 
head(data[, c("restaurant","total_fat")])
```

#### Arrange the data in descending order based on a column
```{r}
# Arranging the data in descending order
data <- data[order(data$total_fat, decreasing = TRUE),]
head(data[, c("restaurant","total_fat")])
```

#### Get the summary of a column -> mean, median, var, SD etc

This function gives us the Minimum value, 1st Quartile value, Median, Mean, 3rd Quartile value, Maximum value from a column in a data frame. 
```{r}
summary(data$total_fat)
```

#### Convert all the values in the column to upper case.

```{r}
data$item <- tolower(data$item)
head(data$item)
```

#### Convert all the values in the column to lower case.
```{r}
data$item <- toupper(data$item)
head(data$item)
```

#### Dropping NAs

Dropping out all the rows with one or more columns as NA
```{r}
data <- openintro::seattlepets
data <- data[complete.cases(data), ]
# Removes the rows with one or more columns having a NA
```


#### groupby( ) and summarize( ) functions usage 
We use ames data to demonstrate these functions. If we want to find the minimum, maximum area of all the houses in a particular neighborhood, we group by on the Neighborhood and compute the minimum and maximum area columns using the summarise function.
```{r}
data <- openintro::ames
data <- data %>% group_by(Neighborhood) %>% summarise(max_area= max(area), min_area=min(area))
head(data, n=10)
```

#### Get the shape of the dataframe in R.

To know the number of rows and columns in a dataframe.
```{r}
data <- openintro::ames
dim(data)
```
First number in the list is the number of rows and the second number in the list is the number of columns in the dataframe


#### Select only top 30 rows or 90 rows, bottom 30 rows
```{r}
data <- openintro::ames
# Selecting the top 30 rows
data <- data[1:30,]
# Selecting the top 90 rows
data <- data[1:90,]
data <- openintro::ames
data <- data[2900:2930,] 
# Select the last 30 rows if your dataframe consists of 2930 rows
```

#### Get the data type of each column in a dataframe.
```{r}
data <- openintro::seattlepets
map(data, class)
```

#### Changing the type of data -> int to char

Change the values in the column from character to integer for all valid values.

Note: Few rows may be marked as NA because of coercion.
```{r warning=FALSE}
data <- openintro::seattlepets
head(data$zip_code)
data$zip_code <- as.numeric(data$zip_code)
head(data$zip_code)
```

#### Rename column names in R

Here we are renaming column name zip_code to zip_code_modified
```{r}
names(data)[names(data) == 'zip_code'] <- 'zip_code_modified'
names(data)
# Zip code has been modified to zip_code_modified
```

#### Find minimum and maximum values in a column in R
```{r}
data <- openintro::ames
min <- min(data$area)
max <- max(data$area)
```
Minimum value of area column
```{r}
min
```
Maximum value of area column
```{r}
max
```

#### Number of unique values in every column in the data frame
In this dataset, we have 13930 unique animal names, 4 different species etc.
```{r}
data <- openintro::seattlepets
data %>% summarize_all(n_distinct)
```

#### Reorder columns in R by column name.
```{r}
authors <- data.frame(
    surname = c("Tukey", "Venables", "Tierney", "Ripley", "McNeil"),
    nationality = c("US", "Australia", "US", "UK", "Australia"),
    retired = c("yes", rep("no", 4)))
authors
#reorder by column name
authors <- authors[, c("retired", "nationality", "surname")]
authors
```

#### Reorder columns by column index.
```{r}
authors <- authors[, c(1,3,2)]
authors
```

#### Remove Duplicates from the dataframe based on one column or multiple columns.

```{r}
# Remove the duplicate rows based on one variable
data <- mtcars
# Have the rows with distinct carb
data_one_var <- distinct(data, carb, .keep_all= TRUE)
# Keep the distinct data based on multiple variables
data_multi <- distinct(data, cyl, vs, .keep_all= TRUE)

```

#### Calculate Mean, Median of a column in the Data Frame
```{r}
mean <- mean(data$carb)
median <- median(data$carb)
```
Mean
```{r}
mean
```
Median
```{r}
median
```


#### Value counts in R. Check how many times a unique variable occurs in like Male - 5, Female -10 in the column name Gender.

Number of rows where cyl = 4 is 11, cyl=6 is 7 etc
```{r}
data %>% count(cyl)
```

Number of rows where carb = 4 is 10, carb=6 is 1 etc
```{r}
data %>% count(carb)
```

#### Convert the index column to a new column

Here we created a new column using the index of the dataframe.
```{r}
data <- cbind(car_name = rownames(data), data)
head(data)
```

#### Add a new row to the dataframe
We can new rows to the dataframe by using ```rbind```
```{r}
new_row_to_add <- data.frame("Volvo 125",22.5,3,120.2,108,2.23,2.89,19.08,1,0,4,3)
names(new_row_to_add) <- c("car_name", "mpg", "cyl", "disp", "hp", "drat", "wt", "qsec", "vs", "am", "gear", "carb")
data <- rbind(data,new_row_to_add)
tail(data)
```

#### Log,Square root, Cube root transformation in R
```{r}
data <- cars
# Taking the log of the speed column
data$log_transformation <- log10(data$speed)
# Taking the square root of the speed column
data$sqrt_transformation <- sqrt(data$speed)
# Taking the cube root of the speed column
data$cube_transformation <-(data$speed)^1/3
head(data)
```

#### Changing the dataframe dimensions from wide to long
Here are the following types on how we can use the ```melt()``` function in R.

Type 1:
Here we create two columns called variable - which represent the subject and value which represents the grade in that subject.
```{r}
df_wide <- data.frame(
  student = c("Krupa", "Goutham", "Sailaja", "Murthy"),
  school = c("St. Joseph's", "Timpany", "St.Joseph's", "Timpany"),
  exploratory_data_analysis = c(10, 100, 1000, 10000),  # eng grades
  probability_and_statistics = c(20, 200, 2000, 20000),  # math grades
  algorithms_for_ds = c(30, 300, 3000, 30000)   # physics grades
)
df_long <- melt(data = df_wide, 
                id.vars = c("student", "school"))
df_long
```
Type 2:
Here if we are interested in the grade of only english and math, we can pass it in the measure.vars parameter.
```{r}
df_long <- melt(data = df_wide, 
                id.vars = "student",
                measure.vars = c("exploratory_data_analysis", "algorithms_for_ds"))
df_long
```


#### Replace NA with a specific value
Here we are replacing NA with None.
```{r}
data <- openintro::seattlepets
data[is.na(data)] <- "None"
```


References:
* https://towardsdatascience.com/data-cleaning-in-r-made-simple-1b77303b0b17
* https://towardsdatascience.com/data-transformation-in-r-288e95438ff9
* https://bookdown.org/aschmi11/RESMHandbook/data-preparation-and-cleaning-in-r.html

