# Parallel coordinate plots cheatsheet

Kechengjie Zhu

```{css, echo=FALSE}
h1 {color: #0146C6}
h2 {color: #3E82FF}
h3 {color: #83AEFF}
```

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
```

***

## Overview
A parallel coordinate plot maps each row in the data table as a line. Packages including **GGally** and **parcoords** help build & improve parallel coordinate plots in R.

***

## Load Packages

```{r message=F,warning=F}
library(GGally)
library(parcoords)
library(d3r)
```

***

## Load Data
Using the **mariokart** data set for illustration.
```{r}
df <- as.data.frame(openintro::mariokart)
```

***

## Basics
```{r}
ggparcoord(data = df,
           column = c(2:7, 9, 11),
           alphaLines = 0.5,) +
  ggtitle("Relations across auction details")
```

### Group by column
Pass to the **groupColumn** argument with a categorical variable representing groups.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond") +
  ggtitle("Relations across auction details grouped")
```

### Grouping Application: Highlight Certain Data Entries
Requires some manipulation on data frame.
```{r}
modified <- df %>%
  mutate(thresh = factor(ifelse(total_pr > 60, "Over 60", "Under 60"))) %>%
  arrange(desc(thresh))
ggparcoord(data = modified,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "thresh") +
  scale_color_manual(values = c("red", "grey")) +
  ggtitle("Highlight sales with total price over $60")
```

### Add data points
Toggle the logical argument **showPoints** to display/hide data points.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           showPoints = TRUE) +
  ggtitle("Relations across auction details with points")
```

### Spline interpolation
Smooth the lines with argument **splineFactor**. Value can be either logical or numeric.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           splineFactor = 7) +
  ggtitle("Smoothed relations across auction details")
```

### Add box plots
Add box plots with **boxplot**.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.2,
           groupColumn = "cond",
           boxplot = TRUE) +
  ggtitle("Relations across auction details with box plots")
```

***

## Scaling methods
Select scaling method with argument **scale**. Default method is **"std"**: subtract mean and divide by standard deviation.

### "robust"
Subtract median and divide by median absolute deviation.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           scale = "robust")
```

### "uniminmax"
Scale so the minimum of the variable is zero, and the maximum is one.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           scale = "uniminmax")
```

### "globalminmax"
No scaling: the range of the graphs is defined by the global minimum and the global maximum.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           scale = "globalminmax")
```

### "center"
Scale using method **"uniminmax"**, and then center each variable at the summary statistic specified by the **scaleSummary** argument.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           scale = "center",
           scaleSummary = "mean")
```

### "centerObs"
Scale using method **"uniminmax"**, and then center each variable at the row number specified by the **centerObsID** argument.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           scale = "centerObs",
           centerObsID = 5)
```

***

## Ordering methods

### "anyClass"
Calculate F-statistics for each class vs. the rest, order variables by their maximum F-statistics.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           order = "anyClass")
```

### "allClass"
Order variables by their overall F-statistic from an ANOVA with **groupColumn** as the explanatory variable.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           order = "allClass")
```

### "skewness"
Order variables by their skewness.
```{r}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond",
           order = "skewness")
```

***

## Make Plots for Each Group with Facets
```{r fig.width=12}
ggparcoord(data = df,
           column = c(2:3, 5:7, 9, 11),
           alphaLines = 0.5,
           groupColumn = "cond") +
  facet_wrap(~ cond) +
  ggtitle("Relations across auction details")
```

***

## Interactive Parallel Coordinate Plots
```{r}
parcoords(df[,c(2:3, 5:7, 9, 11)],
          rownames = F,
          color = list(CcolorBy = "cond"),
          brushMode = "1D-axes",
          reorderable = T,
          queue = T,
          withD3 = T)
```

***

## References
https://www.rdocumentation.org/packages/GGally/versions/1.5.0/topics/ggparcoord
https://www.rdocumentation.org/packages/parcoords/versions/1.0.0/topics/parcoords